---
title: "STAT 301-01 (Sleep Data ANOVA)"
author: "Izzy Mika, Maria Larson, Jiaming Zhang."
date: "2026-04-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Initialize/Setup

```{r}
#install.packages("tidyverse")
#install.packages("readxl")
#install.packages("nlme")
#install.packages("multcomp")
#install.packages("GGally")
#install.packages("factoextra")
#install.packages("ggplot2")
#install.packages("stats")

library(car)
library(tidyverse)
library(readxl)
library(nlme)
library(multcomp)
library(GGally)
library(factoextra)
library(ggplot2)
library(MASS)
library(stats)

# <- read_excel("C:/Users/helix/Documents/MetroState/STAT 301-01 Analysis of Variance and Multivariate Analysis (ANOVA)/Data/___.xlsx")

sleepData <- read_excel("C:/Users/helix/Documents/MetroState/STAT 301-01 Analysis of Variance and Multivariate Analysis (ANOVA)/Project/sleep_health.xlsx")

```

## Project / Data Notes

**Data Changes:**
* No missing values were present.
* Column names were updated to remove spaces for consistency.
* Updated BMI_Cat: combined "Normal Weight" --> "Normal" (same category, based on data description)
* BMI_Cat "Obese" (n = 10 out of 374) was removed due to a small sample size, which could skew results
* Converting Stress_Lvl to Stress_Lvl_Cat by collapsing into 3 levels: 
    * Low (â‰¤3)
    * Medium (4-6)
    * High (>6)

**Limitations:**
* Data are unbalanced across groups. Used lm() with Type II ANOVA of aov() to handle this. 
* Stress Level Category: Low represents approximately -18% of the dataset, while Medium and High are more balanced.
* Other variables (e.g., age, physical activity, occupation) may influence sleep duration but were not included.
* Physical activity was excluded as an IV since it is more appropriate as a continuous variable.

**Visualization Notes:**
* Box plots were avoided since they emphasize medians rather than means.
* Bar charts/interaction plots were used to better represent mean differences.


## Data Analysis

Two-Way ANOVA

364 observations (10 deleted)

1 DV: 
  Sleep Duration
  
2 IVs:
  BMI Category (2 levels: Normal, Overweight)
  Stress Level Category (3 levels: Low, Medium, High)

**Hypotheses:**

*Main Effect of BMI:*
H_0: ÎĽ_Normal = ÎĽ_Overweight 
H_a: ÎĽ_Normal â‰  ÎĽ_Overweight 

*Main Effect of Stress Level Category:*
H_0: ÎĽ_Low = ÎĽ_Medium = ÎĽ_High
H_a: At least one group mean differs

*Interaction:*
H_0: There is no interaction between BMI and stress level category
H_a: There is an interaction between BMI and stress level category


```{r}
# Checking if there is a significant interaction between BMI_Cat and Stress_Lvl_Cat
# --------------------------------------------------------------------------------

# Convert IVs to Factor Variables
sleepData$BMI_Cat <- as.factor(sleepData$BMI_Cat)
sleepData$Stress_Lvl_Cat <- factor(sleepData$Stress_Lvl_Cat , levels = c("Low", "Medium", "High"), ordered = TRUE)  

# Note: Data are unbalanced across groups (especially Stress_Lvl_Cat)
table(sleepData$BMI_Cat)
table(sleepData$Stress_Lvl_Cat)

# Two-Way ANOVA (Unbalanced); Summary
#   **Code: model <- lm(DV~IV1+IV2 + IV1*IV2, data = __)
#           Anova(model, type = "II")
sleepModel <- lm(Sleep_Duration~BMI_Cat+Stress_Lvl_Cat + BMI_Cat*Stress_Lvl_Cat, data = sleepData)
Anova(sleepModel, type = "II")


# Interaction Plot
sleepData$Sleep_Duration <- as.numeric(as.character(sleepData$Sleep_Duration))
interaction.plot(sleepData$BMI_Cat,
                 sleepData$Stress_Lvl_Cat,
                 sleepData$Sleep_Duration,
                 fun=mean,
                 xlab = "BMI Category",
                 ylab = "Mean Sleep Duration (hours)",
                 trace.label = "Stress Level")


```



```{r}

#   Filter to Diff BMI Categories, Test Between Each Stress Lvl Per BMI Cat
#   ------------------------------------------------------------------------

# Filter to Diff BMI Categories 
normWeight <- sleepData %>% filter(BMI_Cat=="Normal")
overweight <- sleepData %>% filter(BMI_Cat=="Overweight")


# What is the test statistic and p-value for comparing the mean sleep duration difference between stress level categories? 
#   Is there any significant difference between stress levels? 
# --------------------------------------------------------------------------------------------------------

# For BMI Category "Normal": 
#   **Code: model <- lm(DV~IV1+IV2 + IV1*IV2, data = __)
#           Anova(model, type = "II ")
normWeightModel <- lm(Sleep_Duration~Stress_Lvl_Cat, data = normWeight)
Anova(normWeightModel, type = "II")

#TukeyHSD(normWeightModel)


# For BMI Category "Overweight": 
#   **Code: model <- lm(DV~IV1+IV2 + IV1*IV2, data = __)
#           Anova(model, type = "II")
overweightModel <- lm(Sleep_Duration~Stress_Lvl_Cat, data = overweight)
Anova(overweightModel, type = "II")


```




```{r}

#   Filter to Diff Stress Level Categories, Test Between Each BMI Cat Per Stress Level 
#   ------------------------------------------------------------------------
# Filter to Diff Stress Level Categories 
lowStress <- sleepData %>% filter(Stress_Lvl_Cat=="Low")
medStress <- sleepData %>% filter(Stress_Lvl_Cat=="Medium")
highStress <- sleepData %>% filter(Stress_Lvl_Cat=="High")


# What is the test statistic and p-value for comparing the mean sleep duration difference between BMI Categories? 
#   Is there any significant difference between BMI Categories? 
# --------------------------------------------------------------------------------------------------------

# Low Stress
#   **Code: t.test(DV~IV, data = ___, alternative = "two.sided")
#       Note: significant difference == "two.sided"
t.test(Sleep_Duration~BMI_Cat, data = lowStress, alternative = "two.sided")


# Medium Stress
t.test(Sleep_Duration~BMI_Cat, data = medStress, alternative = "two.sided")


# High Stress
t.test(Sleep_Duration~BMI_Cat, data = highStress, alternative = "two.sided")



```



```{r}

# Bar Chart - Mean Sleep Duration by Stress Level and BMI Category 

summary_data <- sleepData %>%
  group_by(Stress_Lvl_Cat, BMI_Cat) %>%
  summarise(mean_sleep = mean(Sleep_Duration), se = sd(Sleep_Duration)/sqrt(n()))

summary_data$Stress_Lvl_Cat <- factor(summary_data$Stress_Lvl_Cat, 
                                      levels = c("Low","Medium","High"), 
                                      labels = c("Low (â‰¤3)", "Medium (4-6)", "High (>6)"))

p <- ggplot(summary_data, aes(x = Stress_Lvl_Cat, y = mean_sleep, fill = BMI_Cat))

p <- p + geom_bar(stat = "identity", position = "dodge")

p <- p + geom_errorbar(aes(ymin = mean_sleep - se, ymax = mean_sleep + se), 
                       position = position_dodge(0.9), width = 0.25)

p <- p + labs(title = "Mean Sleep Duration by Stress Level and BMI Category", 
              x = "Stress Level Category", 
              y = "Mean Sleep Duration (hours)", 
              fill = "BMI Category")

p <- p + theme_minimal()

p

```

